---
title: 'Automatic or Manual Transmission: Which Has Better Fuel Economy?'
author: "S. M. Kaiser"
date: "June 15, 2015"
output: pdf_document
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(car)
library(knitr)
library(ggplot2)
library(stargazer)
library(xtable)
#opts_chunk$set(echo=FALSE)
theme_set(theme_grey(base_size = 8))
data(mtcars)
mpg <- mtcars$mpg
am <- factor(mtcars$am, levels=c(0,1), labels=c("Automatic", "Manual"), ordered = TRUE)
d <- data.frame(mpg, am, cyl=mtcars$cyl)

```
# Executive Summary
Ignoring other variables, cars with manual transmissions are more fuel-efficient than those with automatics. On average, cars with manual transmissions achieve 24.4 miles per gallon, while automatics achieve 17.1 miles per gallon. 

However, looking at all the data available, there are other factors to consider along with the 
transmission type. 

Manual transmissions happen to be more fuel-efficient than automatics due to these factors:

- Cars with fewer cylinders are more likely to be manual
- Cars with smaller displacement are more likely to be manual
- Cars with lower weight are more likely to be manual

In conclusion, transmission type is related to fuel economy, but not as much as the three factors above.

# Exploratory Analysis

See [Coefficients](#coefficients) for an explanation of each variable.

A boxplot shows that manual transmissions are generally more fuel-efficient than automatics. However, there may be other factors that influence fuel economy.

```{r echo=FALSE, fig.height=2, fig.width=3, fig.cap = "Boxplot of fuel economy by transmission type"}
ggplot(d)  + geom_boxplot(aes(y=mpg, x=am)) + ylab("Fuel Economy (MPG)") + xlab("Transmission Type")
```

# Model selection
Fitting a [simple model](#simplemodel) with `am` as the predictor and `mpg` as the outcome, there appears to be a slightly significant relationship between `am` and `mpg`, with a correlation of 0.6.

Using VIF, we can identify the other variables that have a greater impact on `mpg` besides `am`. The top three by a significant margin are displacement (`disp`), cylinders (`cyl`) and weight (`wt`).
```{r echo=FALSE}
model2 <- lm(mpg ~ ., data=mtcars)
```

```{r results="asis", echo=FALSE}
stargazer(vif(model2), header = FALSE, title = "VIF for each predictor")
```


# Appendix

## Coefficients {#coefficients}
These are the available variables in the data. For more information in R, see `?mtcars`.

|Variable|Description
|---|--------------------
|mpg|Miles/(US) gallon
|cyl|Number of cylinders
|disp|Displacement (cu.in.)
|hp|Gross horsepower
|drat|Rear axle ratio
|wt|Weight (lb/1000)
|qsec|1/4 mile time in seconds
|vs|"V" or straight cylinder layout
|am|Transmission (0 = automatic, 1 = manual)
|gear|Number of forward gears
|carb|Number of carburetors

## Pairs Plot
```{r echo=FALSE}
pairs(mtcars, panel = panel.smooth)
```

## Linear Regression

### Simple Model (mpg ~ am) {#simplemodel}

```{r results='asis', echo=FALSE}
model1 <- lm(mpg ~ am, data=mtcars)
stargazer(model1, header = FALSE, title = "Linear Model with transmission type (am) as the predictor, and MPG as the outcome")
```
```{r echo=FALSE, fig.width=4, fig.height=3}
ggplot(d,aes(y=mpg, x=am)) + geom_point() + geom_smooth(method="lm", aes(group=1)) + ylab("Fuel Economy (MPG)") + xlab("Transmission Type")
```

### R Plots of Simple Model

```{r echo=FALSE}
par(mar=c(2,2,2,2))
par(mfrow=c(2,2))
plot(model1)
```

### Residual Plot of Selected Model {#residualplot}
```{r echo=FALSE}
model3 <- lm(mpg ~ . - cyl - disp - wt, data=mtcars)
plot(fitted(model3), resid(model3))
abline(h=0)

```

## Sources
- [Variance Inflation Factor](https://github.com/mthaugs/Data-Science-Diagnostics/blob/master/VIF.pdf)
- [Statistical Modeling: The Two Cultures](http://projecteuclid.org/download/pdf_1/euclid.ss/1009213726)
- Hlavac, Marek (2014). stargazer: LaTeX code and ASCII text for well-formatted regression and summary statistics tables. http://CRAN.R-project.org/package=stargazer
